Reversible Regime Change: Climate-driven phytoplankton community shifts in the Cariaco Basin, Venezuela

### 2025-12-15 text-backup.tex
Pre-Edits for Review Response

>>>>

%OLDABSTRACT

- The CARIACO phytoplankton community time series is characterized by two distinct temporal clusters separated by a shift in 2004 from high to low upwelling and by a return to high upwelling conditions in 2014.
- No significant differences are observed in diversity metrics between the two clusters, with the exception of a lower turnover and a higher community evenness during low upwelling conditions.
- Gradient forest analysis identifies the AMO climate index as the strongest predictor of community change, highlighting the influence of large-scale climatic oscillations on this tropical coastal ecosystem

Phytoplankton communities are integral to oceanic biogeochemical cycles and are sensitive indicators of climate-driven environmental variability.
Long-term time series can capture this variability, allowing us to unravel the effects of environmental change on local communities. This study investigates changes in the phytoplankton community of the Cariaco Basin, off the coast of Venezuela, from 1995 to 2017, using monthly water column observations integrated with climate indices and meteorological data. To investigate impacts of larger-scale climate variability, we considered the Atlantic Multidecadal Oscillation (AMO) index and the Multivariate El Niño Southern Oscillation index (MEI v.2), both of which have been linked to variations in local weather and upwelling in the Cariaco Basin.
Cluster analysis identified two distinct community states separated by a shift from high to low upwelling conditions in 2004. In 2014, community and environmental conditions returned to a state similar to that before the shift in 2004. In contrast, community diversity declined between 1998 and 2009, independent of changes in the upwelling regime.
To quantify the effects of climate and bottom-up drivers on phytoplankton community changes, we employed a gradient forest analysis and found that the AMO index was the strongest predictor of change, closely followed by in situ variables related to upwelling (sea surface temperature and nitrate concentration) and the MEI v.2. Our study emphasizes the impact of long-term climatic oscillations, such as the AMO, on the phytoplankton community of a tropical coastal ecosystem through modulation of the upwelling system and suggests a possible cyclic (if not resilient) behavior of this tropical marine ecosystem under global change.
